Some examples in extended logic counting classes

### 2016-09-16 QSO/logics.tex
15 sep

We define an operator which extends the Least Fixed Point logic to counting. Recall that a fixed point operator is defined by a positive formula $\varphi(x_1,\ldots,x_k,R)$ where $R$ is a predicate or arity $k$. The operator $T_{\varphi}:2^{A^k} \to 2^{A^k}$ is defined as $T_{\varphi}(X) = \{(a_1,\ldots,a_k)\mid (\A,X)\models \varphi(a_1,\ldots,a_k,R) \}$, for each $X\subseteq A^k$ and $X^*\subseteq A^k$ is a fixed point of $T_{\varphi}$ if $T_{\varphi}(X) = X$. Then, the formula $[\textbf{lfp}_{(x_1,\ldots,x_k),R}\varphi(x_1,\ldots,x_k,R)](y_1,\ldots,y_k)$ is evaluated at the least fixed point of $T_{\varphi}$.

We introduce the operator $[\textbf{alfp}_{\varphi(x_1,\ldots,x_k,R)}\alpha(u_1,\ldots,u_k,R,\pi)](y_1,\ldots,y_k)$ which uses a FO-formula $\varphi$ as well as a QSO-formula $\alpha$ and takes as input a placeholder function $\pi$, which is interpreted as a constant. The partial function $\zeta_{\varphi,\alpha}:2^{A^k} \to \nat$ is such that  $\zeta_{\varphi,\alpha}(T_{\varphi}(X)) = \sem{\alpha'}(\A,R)$ where $\alpha'$ is obtained by replacing every instance of $\pi(\bar{z})$ by $\zeta_{\varphi,\alpha}(X)$, and $\zeta_{\varphi,\alpha}(\emptyset) = 0$. Then the operator is evaluated as $$\sem{[\textbf{alfp}_{\varphi(x_1,\ldots,x_k,R)}\alpha(u_1,\ldots,u_k,R,\pi)](y_1,\ldots,y_k)}(y_1,\ldots,y_k) = \zeta_{\varphi,\alpha}(X^*)$$, where $X^*$ is the least fixed point of $T_{\varphi}$.

>>>>

We define an operator which extends the Least Fixed Point logic to counting. Recall that a fixed point operator is defined by a formula $\varphi(x_1,\ldots,x_k,R)$ which is positive on $R$ where $R$ is a predicate or arity $k$. The operator $T_{\varphi}:2^{A^k} \to 2^{A^k}$ is defined as $T_{\varphi}(X) = \{(a_1,\ldots,a_k)\mid (\A,X)\models \varphi(a_1,\ldots,a_k,R) \}$, for each $X\subseteq A^k$ and $X^*\subseteq A^k$ is a fixed point of $T_{\varphi}$ if $T_{\varphi}(X^*) = X^*$. Then, the formula $[\textbf{lfp}\,\varphi(x_1,\ldots,x_k,R)](y_1,\ldots,y_k)$ is evaluated at the least fixed point of $T_{\varphi}$.

We extend QSO with the operator $[\textbf{alfp}_{\varphi(y_1,\ldots,y_k,R)}\alpha(x_1,\ldots,x_{\ell},R,\pi)](x_1,\ldots,x_{\ell})$. It is defined by a FO-formula $\varphi$ and a QFO-formula $\alpha$ which mentions a placeholder function $\pi$ with domain $\fv^{\ell}$.

To formally characterize its semantics we define the partial function $\zeta_{\varphi,\alpha}:2^{A^k} \to (A^{\ell}\times\nat)$ for each $\varphi$ and $\alpha$ as follows. Let $\A$ be a structure. For $R = \emptyset$, it holds that $\zeta_{\varphi,\alpha}[\emptyset] = \sem{\alpha\mid_{\pi(\bar{u})\to 0}}$. For each $R \subseteq A^k$ such that $T_{\varphi}(R) \neq R$, let $\beta$ be a virtual $\qfo$ formula such that $\sem{\beta}(\A,v) = \zeta_{\varphi,\alpha}[R](v)$ for each first order assignment $v$, and it holds that $\zeta_{\varphi,\alpha}[T_{\varphi}(R)] = \sem{\alpha\mid_{\pi\to\beta}}$.

For a given first order asignment $v$, the operator is evaluated as:
$$
[\textbf{alfp}_{\varphi(y_1,\ldots,y_k,R)}\alpha(x_1,\ldots,x_{\ell},R,\pi)](\A,v) = \zeta_{\varphi,\alpha}[X^*](v),
$$
where $X^*$ is the least fixed point of $\varphi$.
